Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-05 TEBenchmarkPaper/article.tex
Clarification

We chose TIP for our ground truth since it satisfies the desiderata of
section~\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes higher-order functions and
inductive datatypes, which we are interested in; plus it is accompanied by
tooling to convert this format into a variety of languages, including Haskell
and Isabelle.

We use TIP version 0.2 which contains 343 problems, each stating a single
theorem and together defining a total of 618 datatypes and 1498 functions. Most
of these are duplicates, since each problem (re\nobreakdash-)defines all of the
datatypes and functions it involves. To avoid ambiguity we prefix the names of
all definitions with the filename it came from, and collect them into one big set.

TIP datatypes can have several ``constructors'' and ``destructors''. For example
the type of lists from figure~\ref{figure:list_theory} can be defined in TIP as
follows:

\begin{verbatim}
(declare-datatypes (a) ((List (Nil) (Cons (head a) (tail (List a))))))
\end{verbatim}

This defines a type \texttt{List} parameterised by a type variable \texttt{a}
(the type of the list's elements), with constructors \texttt{Nil} and
\texttt{Cons}, and destructors \texttt{head} and \texttt{tail}. Target languages
(e.g. Haskell and Isabelle) handle constructors and destructors differently, so
we avoid exploring them directly. Rather, we $\eta$-expand each constructor and
destructor to create new function definitions, and rewrite the TIP theorems to
reference these expanded forms instead.

TIP has ``native'' support for booleans and integers, allowing functions like
\texttt{+} to be used without an accompanying definition. To ensure consistency
in the translations, we replace all occurrences of these ``native'' expressions
with standard definitions using the same \texttt{declare-datatypes} and
\texttt{define-fun} mechanisms as the problem set.

This gives us a total of 3598 definitions, including those from the problem set
and our automatically-generated additions. Removing $\alpha$-equivalent
duplicates leaves 269 definitions, and we chose to only sample from those 182
which occur in at least one theorem statement (this removes ambiguity about
which \emph{definitions}, including our generated ones, are ``interesting'' and
which are just ``implementation details'').

TIP includes a tool to translate its definitions into Haskell code, including
suitable comparison functions and random data generators. We translated all 269
definitions into a single Haskell package, compiled it with GHC and extracted
the function definitions with our plugin. Each benchmark run imported this
entire package, but the signature given to QuickSpec only included functions
chosen for the current sample.

@@

We applied our sampling methodology to the 182 definitions which appear in TIP
theorems, for even sample sizes up to 20. We chose 30 samples for each size, but
due to the requirement for samples to admit at least one theorem, only 29 unique
samples were obtained for size 2. A timeout of 5 minutes was imposed, since
preliminary experiments showed that most QuickSpec runs either finish before
this time, or still do not finish after an hour. The running times are shown in
figure~\ref{figure:quickspec_runtimes}.

The only errors we encountered were timeouts (shown in red), and these occurred
for all sample sizes. The running times cluster, with many close to zero and
many exceeding the timeout, but with few in between. There is a clear trend
towards longer times and more timeouts as the sample size increases, such that
most samples of size 20 timed out.

Most runs finished quickly and generated few conjectures, with the most being
116 conjectures from a sample of size 16.

@@

We use a binomial distribution to model precision and recall, as if each
conjecture is a ball drawn from an urn of wanted/unwanted results (in the case
of precision) or found/missed ground truths (in the case of recall). The mean of
these distributions estimates the overall proportion of wanted or found
conjectures, and we use the standard deviation to determine how well this mean
fits the data.

>>>>

We chose TIP for our ground truth since it satisfies the desiderata of
section~\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes higher-order functions and
inductive datatypes, which we are interested in; plus it is accompanied by
tooling to convert this format into a variety of languages, including Haskell
and Isabelle which are used by existing MTE tools.

We use TIP version 0.2 which contains 343 problems, each stating a single
theorem and together defining a total of 618 datatypes and 1498 functions. Most
of these are duplicates, since each problem (re\nobreakdash-)defines all of the
datatypes and functions it involves.

TIP datatypes can have several ``constructors'' (introduction forms) and
``destructors'' (elimination forms; field accessors). For example the type of
lists from figure~\ref{figure:list_theory} can be defined in the TIP format as
follows:

\begin{verbatim}
(declare-datatypes
  (a)                       ;; Type variable (element type)
  ((List                    ;; Type name
     (Nil)                  ;; Constructor (nullary)
     (Cons                  ;; Constructor (binary)
       (head a)             ;; Field name and type
       (tail (List a))))))  ;; Field name and type
\end{verbatim}

Our target languages (Haskell and Isabelle) differ in the way they handle
constructors and destructors, which complicates comparisons. To avoid this, we
generate a new function for each constructor (via $\eta$-expansion) and
destructor (via pattern-matching) of the following form:

\begin{verbatim}
(define-fun
  (par (a)                   ;; Type variable
    (constructor-Cons        ;; Function name
      ((x a) (xs (List a)))  ;; Argument names and types
      (List a)               ;; Return type
      (as                    ;; Type annotation
        (Cons x xs)          ;; Return value
        (List a)))))         ;; Return type

(define-fun
  (par (a)                        ;; Type variable
    (destructor-head              ;; Function name
      ((xs (List a)))             ;; Argument name and type
      a                           ;; Return type
      (match xs                   ;; Pattern-match
        (case (Cons h t) h)))))   ;; Return relevant field
\end{verbatim}

\begin{sloppypar}
  We rewrite the TIP theorems (our ground truth) to reference these expanded
  forms instead of the raw constructors and destructors, and include these
  functions in our samples in lieu of the raw expressions. Note that these
  destructor wrappers are \emph{partial} functions (e.g.
  \texttt{destructor-head} and \texttt{destructor-tail} are undefined for the
  input \texttt{Nil}), which complicates their translation to proof assistants
  like Isabelle.
\end{sloppypar}

Another complication is TIP's ``native'' support for booleans and integers,
which allows numerals and symbols like \texttt{+} to appear without any
accompanying definition. To ensure consistency in the translations, we replace
all occurrences of such expressions with standard definitions written with the
``user-level'' \texttt{declare-datatypes} and \texttt{define-fun} mechanisms.

When we add all of these generated types and functions to those in TIP, we get a
total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
and we choose to only sample from those 182 which occur in at least one theorem
statement (this removes ambiguity about which \emph{definitions} count as
``interesting'' and which are just ``implementation details'' for other
definitions).

TIP includes a tool to translate its definitions into Haskell and Isabelle,
including suitable comparison functions and random data generators. We translate
all 269 unique definitions into a single module/theory which is imported on each
run of the MTE tools (although only those functions which appear in the current
sample are actually explored).

@@

We applied our sampling methodology to the 182 definitions which appear in TIP
theorems, for even sample sizes up to 20. We chose 30 samples for each size,
although one sample of size 2 was discarded as it was a duplicate. A timeout of
5 minutes was imposed, since preliminary experiments showed most QuickSpec runs
to either finish before this time, or still be running after an hour. The
running times are shown in figure~\ref{figure:quickspec_runtimes}.

The only errors we encountered were timeouts (shown in red), which occurred for
all sample sizes but are more common for larger samples (over half the runs for
sample size 20 timed out). Runs which did not time out mostly finished within 30
seconds, generating few or no conjectures; those taking longer generated more
conjectures, with the most being 116 conjectures from a sample of size 16.

@@

We use a binomial distribution to model precision and recall, as if each
conjecture is drawn from a pool of wanted/unwanted results (in the case of
precision) or found/missed ground truths (in the case of recall), with the mean
of these distributions estimating the proportions of each pool's split.
